---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
## First read the data, then convert the date into class date, and finaly print the first few elements
```{r}
## Read the data
unzip("activity.zip" )
data<- read.csv("./activity.csv",  sep=",", header = TRUE)

## convert the date column into class date
data$date <- as.Date(data$date, format = "%Y-%m-%d")

## print the first few elements
head(data)

```


## What is mean total number of steps taken per day?
## we use the aggregate function to find the mean for each day, then hist is used to plot the histogram. the mean and median are printed at the end
```{r}
## get the mean across each day
meandata <- aggregate(data$steps, list(data$date), mean, na.action = na.omit, na.rm=TRUE)

## plot the histogram for daily steps
hist(meandata$x , col = "red", xlab = "Mean Steps per Day", main = "Mean Steps")

## mean number of steps
mean(meandata$x, na.rm = TRUE)

## median number of steps
median(meandata$x, na.rm = TRUE)

```


## What is the average daily activity pattern?
## again we use the aggregate function, and the plot function to plot the time series. finally the maximum interval is found
```{r}
## get the mean across intervals
dailydata <- aggregate(data$steps, list(data$interval), mean, na.action = na.omit, na.rm=TRUE)

## plot the average time series
plot(dailydata$x,type="l" , ylab = "Daily Pattern",xlab = "Time Interval", main = "Daily Activity Pattern")

## The interval with the maximum number of steps is at 
which.max(dailydata$x)
dailydata$Group.1[which.max(dailydata$x)]
```


## Imputing missing values
## we replace each na value with the correspinding interval mean, the mean and median are printed at the end. The mean is still the same but the median is different. The histogram shows a larger peak at the mean value.
```{r}

sum(is.na(data$steps))

## copy data into a new data frame
datanona <- data

## loop to replace each na value with the mean of that interval
for(i in 1:nrow(datanona)){
  if (is.na(datanona[i,1])){
    datanona[i,1] = dailydata[which(dailydata$Group.1 == datanona[i,3]),]$x
  }
  
}

## confirm all na values are changed
sum(is.na(datanona$steps))

## get the mean across days
meandatanona <- aggregate(datanona$steps, list(datanona$date), mean, na.action = na.omit, na.rm=TRUE)

## plot the histogram
hist(meandatanona$x , col = "red", xlab = "Mean Steps per Day", main = "Mean Steps")

## The mean number of steps
mean(meandatanona$x, na.rm = TRUE)

## median number of steps
median(meandatanona$x, na.rm = TRUE)

## The mean value is the same as before but the median is different and there is a slight change in the histogram
```




## Are there differences in activity patterns between weekdays and weekends?
## the weekdays have a larger peak at an earlier time in the morning .
```{r}
## vector of weekend days
weekend = c("Saturday","Sunday")

## add a factor column indicating whether it is a weekday or weekend
data$day <- factor(weekdays(data$date) %in% weekend, labels = c("Weekday","weekend"))

## get the mean for each interval for each category, weekday or weekend
meandataweekde <- aggregate(datanona$steps, list(data$day,data$interval), mean, na.action = na.omit, na.rm=TRUE)
meandataweekde$Group.1 <- factor(meandataweekde$Group.1)
library(lattice)

## plot the time series for weekdays and weekends
xyplot(   x ~ seq(1,length(meandataweekde$x),1) | Group.1, data = meandataweekde, type="l", ylab = "Number of Steps", xlab = "Interval",  layout = c(1,2))
```

